---
# date: "`r Sys.Date()`"
output: github_document
header-includes:
  - \usepackage{amsfonts,amsmath,amsthm,amsbsy,amssymb,dsfont,bm,mathtools,mathalfa}
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  tidy = TRUE, 
  tidy.opts = list(comment = FALSE),
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
library(epca)
```

# Exploratory Principal Component Analysis

<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

`epca` is an R package for comprehending any data matrix that contains *low-rank* and *sparse* underlying signals of interest. 
The package currently features two key tools:

* `sca` for **s**parse principal **c**omponent **a**nalysis.
* `sma` for **s**parse **m**atrix **a**pproximation, a two-way data analysis for simultaneously row and column dimensionality reductions.

## Installation

<!-- You can install the released version of epca from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("epca") -->
<!-- ``` -->

<!-- or the development version from [GitHub](https://github.com/) with: -->

`epca` is not yet on CRAN. You could install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("fchen365/epca")
```
## Example

Here is an example with simulated data. First, simulate a rank-5 data matrix with some additive Gaussian noise.
```{r simu, eval=F}
## simulate a rank-5 data matrix with some additive Gaussian noise 
n <- 300
p <- 50
k <- 5 ## rank
Z <- shrinkage(polar(matrix(runif(n * k), n, k)), sqrt(n * k))
B <- diag(5) * 3
Y <- shrinkage(polar(matrix(runif(p * k), p, k)), sqrt(p * k))
E <- matrix(rnorm(n * p, sd = .01), n, p)
X <- scale(Z %*% B %*% t(Y) + E)
```

Now, perform the sparse PCA.
```{r sca, eval=F}
## perform sparse PCA
sca(X, k)
```

Similarly, we can do sparse matrix decomposition.
```{r sma, eval=F}
## perform sparse matrix approximation
sma(X, k)
```


## Reference

Chen F and Rohe K, "A New Basis for Sparse PCA."

