---
title: "Explore multivariate data with `epca`"
author: "Fan Chen (fan.chen@wisc.edu)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Explore multivariate data with `epca`}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      tidy = TRUE, 
                      tidy.opts = list(comment = FALSE))
library(epca)
library(Matrix)
library(tidyverse)
```

This document provides three examples of how to use `epca` to explore your data.

## Example 1: simulated data

The first example shows how to call the two key functions: `sca` and `sma`.
Let's simulate a rank-5 data matrix with some additive Gaussian noise.
```{r simu}
## simulate a rank-5 data matrix with some additive Gaussian noise 
n <- 300
p <- 50
k <- 5 ## rank
Z <- shrinkage(svd(matrix(runif(n * k), n, k))$u, gamma = sqrt(n))
B <- diag(5) * 3
Y <- shrinkage(svd(matrix(runif(p * k), p, k))$u, gamma = sqrt(p))
E <- matrix(rnorm(n * p, sd = .01), n, p)
X <- scale(Z %*% B %*% t(Y) + E)
```
In the above code, matrices `Z` and `Y` are rank-5 and sparse. The `shrinkage()` function performs a soft-thresholding by default.

Now, apply `sca` to find 5 sparse PCs. 
```{r sca}
## perform sparse PCA
s.sca <- sca(X, k = 5)
s.sca
```

Similarly, we can do sparse matrix decomposition.
```{r sma}
## perform sparse matrix approximation
s.sma <- sma(X, k = 5)
s.sma
```

## Example 2: `pitprops` data 
For the second example example, we use the `pitprops` data.

We apply the `sca` function to find `k = 3` sparse PCs with sparsity parameter `gamma = 4.5`. Here, the sparsity parameter (`gamma`) controls the L1 norm of the returned PC loadings. The default of `gamma` (if absent) is `sqrt(p * k)`, where `p` is the number of original variables. 
```{r 3pc}
## find 3 sparse PCs
data("pitprops", package = "epca")
s.sca <- sca(pitprops, k = 3, gamma = 4.5)
print(s.sca, verbose = TRUE)
```
Here, option `verbose = TRUE` prints, for each PC, the original variable with non-zero loadings.

The followings find 6 sparse PCs with `gamma = 6`. Note that the sparsity parameter for 
```{r 6pc}
## find 6 sparse PCs
s.sca <- sca(pitprops, 6, gamma = 6)
print(s.sca, verbose = TRUE)
```

## Example 3: single-cell RNA-seq data

```{r import results, echo=FALSE}
load("scrnaseq.rda")
```

This example shows a large-scale application of sparse PCA to a single-cell RNA-seq data. 
For this example, we use the human/mouse pancreas single-cell RNA-seq data from Baron et al. (2017). 


Fe used the single-cell RNA-seq data with the `scRNAseq` package. We removed the genes that do not have any variation across samples (i.e., zero standard deviation) and the cell types that contain fewer than 100 cells. This resulted in a sparse data matrix `pancreas` of 17499 genes (rows) and 8451 cells (columns) across nine cell types.
```{r import scRNA-seq data, eval=FALSE}
# library(scRNAseq)
dat <- BaronPancreasData('human')
# dim(dat) ## 20125  8569
gene.select <- !!apply(counts(dat), 1, sd) ## remove non-variance gene
label.select <- colData(dat) %>% 
  data.frame() %>% 
  dplyr::count(label) %>% 
  filter(n > 100) 
#   label                  n
# 1 acinar               958
# 2 activated_stellate   284
# 3 alpha               2326
# 4 beta                2525
# 5 delta                601
# 6 ductal              1077
# 7 endothelial          252
# 8 gamma                255
# 9 quiescent_stellate   173
dat1 <- dat[gene.select, colData(dat)$label %in% label.select$label]
```

For SCA, we use the expression count matrix (`count`) as the input, where `count[i,j]` is the expression level of gene j in cell i, with 10.8\% being non-zero. 
```{r extract count matrix, eval=FALSE}
count <- counts(dat1)
# dim(count) ## 17499  8451
# length(count@i) / length(count) ## %(nnz)
## 10.80605% non-zeros
```

The dataset contains labels for each cell.
```{r extract cell label, eval=FALSE}
label <- setNames(factor(dat1$label), colnames(dat1))
```

Next, We applied `sca` to the transpose of `count` to find `k = 9` sparse gene PCs. Aiming for a small number of genes (i.e., non-zero loadings) in individual PCs, we set the sparsity parameter to `gamma = log(pk)`, which is approximately 12. 
<!-- The algorithm took 24 iterations and about 5 minutes on a single processor (3.3GHz). -->
```{r apply sca to scRNA-seq, eval=FALSE}
scar <- sca(t(count), k = 9, gamma = 12,
             center = F, scale = F, 
             epsilon = 1e-3)
```

We can exam the number of original genes included by each gene PC. 
```{r number of non-zeros}
n.gene <- apply(!!scar$loadings, 2, sum)
n.gene
```
Each gene PC uses a handful of original genes. 


We can plot the component scores of the nine PCs, with `dplyr` and `ggplot2` packages. Each panel displays one of nine cell types with the names of cell types and the number of cells reported on the top strips. For each cell type, a box depicts the component scores for nine sparse gene PCs.
```{r plot, fig.width=6, fig.height = 6, fig.cap="Scores of sparse gene principal components (PCs) stratified by cell types."}
scar$scores %>%
  reshape2::melt(varnames = c("cell", "PC"), 
                 value.name = "scores") %>% 
  mutate(PC = factor(PC), label = label[cell]) %>%
  ggplot(aes(PC, scores / 1000, fill = PC)) +
  geom_boxplot(color = "grey30", outlier.shape = NA, 
               show.legend = FALSE) + 
  labs(x = "gene PC", y = bquote("scores ("~10^3~")")) + 
  scale_x_discrete(labels = 1:9) + 
  facet_wrap(~ label, nrow = 3) + 
  scale_fill_brewer(palette = "Set3") +
  theme_classic() 
```

We observed that most of the gene PCs consist of one or a handful of genes, yet the component scores showed that these PCs distinguish different cell types effectively . For example, the PC 2 consists of only one gene (named SST), and the expression of the gene marks the "delta" cells among others. This result highlights power of scRNA-seq in capture cell-type specific information and suggests the applicability of our methods to biological data.
